From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-30 upi.tex
move wavefront deeper in section 3.  Split the two non-examples (easier to explain).  Add more words.

@@

As two simple examples of non-univalent fibrations consider
$P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P = \lambda\_.\mathbb{2}$
and $Q : \mathbb{2} \rightarrow \mathbb{U}$ defined as
$Q = \lambda \_. \mathbb{1}$ illustrated below:

\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{1}$};
  \draw[fill] (0,-5) circle [radius=0.025];
  \node[below] at (0,-5) {$\star$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $P(\star) = \mathbb{2}$};
  \draw[fill] (-0.2,-2.7) circle [radius=0.025];
  \draw[fill] (0.2,-2.7) circle [radius=0.025];
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$\fc$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$\tc$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $Q(\fc)$ = Fiber $Q(\tc)$ = $\mathbb{1}$};
  \draw[fill] (0,-2.7) circle [radius=0.025];
\end{tikzpicture}
\end{tabular}

\medskip\noindent The identity function and boolean negation induce two distinct
equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
P(\star)$. As there is only one path $\star \equiv \star$ in the base space, the
fibration $P$ is not univalent. Similarly, there is one equivalence
$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but this certainly does not
induce a path $\fc \equiv \tc$. As the examples illustrate there is a delicate
balance required for the fibration to be univalent; otherwise some distinct
equivalences would be identified (like for $P$) or distinct points in the base
space would be identified (like for $Q$).

>>>>

As is often the case when confronted with new definitions, it is
instructive to disect \emph{non-examples}, as a means to understand
what ``goes wrong''.  As a first example of a non-univalent fibration,
consider $P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P =
\lambda\_.\mathbb{2}$.

\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{1}$};
  \draw[fill] (0,-5) circle [radius=0.025];
  \node[below] at (0,-5) {$\star$};
  \draw (-1.3,-2.7) ellipse (0.5cm and 1cm);
  \draw[fill] (-1.5,-2.7) circle [radius=0.025];
  \draw[fill] (-1.1,-2.7) circle [radius=0.025];
  \node[left] at (-1.8,-2.7) {Fiber $P(\star) = \mathbb{2}$};
  \draw (1.3,-2.7) ellipse (0.5cm and 1cm);
  \draw[fill] (1.1,-2.7) circle [radius=0.025];
  \draw[fill] (1.5,-2.7) circle [radius=0.025];
  \node[right] at (1.8,-2.7) {Fiber $P(\star) = \mathbb{2}$};
  \node[below,cyan,dashed,thick] at (0,-5) {$\equiv$};
  \draw[->,red,dashed,ultra thick] (-0.8,-2.3) to [out=45, in=135] (0.8,-2.3);
  \draw[->,red,dashed,ultra thick] (0.8,-3.1) to [out=-135, in=-45] (-0.8,-3.1);
  \node[red,ultra thick] at (0,-2.7) {$\simeq$};
\end{tikzpicture}

\medskip\noindent There is actually \emph{a single fiber}, but
the identity function and boolean negation induce \emph{two distinct
equivalences} of the space $P(\star) = \mathbb{2}$ with itself.
In other words, $\mathbb{2} \simeq \mathbb{2}$ has two distinct inhabitants.
As there is only one path $\star \equiv \star$ in the base space, the
fibration $P$ is not univalent.

Another example is given by $Q : \mathbb{2} \rightarrow \mathbb{U}$ defined as
$Q = \lambda \_. \mathbb{1}$ illustrated below

\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$\fc$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$\tc$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $Q(\fc)$ = Fiber $Q(\tc)$ = $\mathbb{1}$};
  \draw[fill] (0,-2.7) circle [radius=0.025];
\end{tikzpicture}

The situation here is different.  There is one equivalence
$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but it does not come
from a path (such as $\fc \equiv \tc$, which does not exist).  And that
equivalence does not ``induce'' such a path either.

As the examples illustrate there is a delicate
balance required for the fibration to be univalent; otherwise some distinct
equivalences would be identified (like for $P$) or distinct points in the base
space would be identified (like for $Q$).

@@
